---
title: "ds202_hw4"
author: "Aashwin Lamsal"
date: "10/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Homework 4: Part 1
Download the Ames specific liquor sales data to my local files

```{r}
if (file.exists("alcohol.rds")) {
  data <- readRDS("alcohol.rds")
} else {
  url <- "https://github.com/lbchu/ds202_labs/raw/master/ames.csv"
  data <- readr::read_csv(url)
  saveRDS(data, file="alcohol.rds")
} 
```

## Part 2: Data Cleaning

The following is a snapshot of the data, and judging by this, I'm going to be changing the variable names of the ones that have a space separating the 2 words that make up its respective name.

```{r}
head(data)
#print(data, width=Inf) commented out because it made the HTML file too lengthy #many of the columns have spaces between their respective column names, so using the substitute function, I'm going to replace the spaces with underscores
names(data)

names(data) <- sub(" ", "_", names(data)) #since sub replaces the first instance of the space, I ran it twice so that it can catch all the spaces, especially in the longer variable names
names(data) <- sub(" ", "_", names(data))

names(data) # ran again to check if the changes had occurred

str(data)
```

Next we extract the geographic lattitude and longitude from "Store_Location".

```{r}
library(tidyverse)
newData <- data
point_split = strsplit(data$Store_Location, split = " ") #this line splits the Store Location at each instance of a space
newData = transform(newData, Point = sapply(point_split, "[", 1), lattitude= sapply(point_split, "[", 2), longitude = sapply(point_split, "[", 3))


newData <- newData[-8]#removing the "Store_Location" column since we extracted the latt. and long. already.
newData <- newData[-24] #removing the "Point" column, since all it had as its entries were the word "POINT"
#Next, I have to remove the left parenthesis and right parenthesis for lattitude and longitude respectively.

newData$lattitude <- gsub("\\(", "", newData$lattitude) #left parenthesis removed
newData$longitude <- gsub("\\)", "", newData$longitude) #right parenthesis removed

#Replacing the INVOICE.Item_NUMBER "INV-" characters so all that remains is just the numbers, for type transformation later
newData$Invoice.Item_Number <- gsub("INV-", "", newData$Invoice.Item_Number)
#Now I choose 5 variables that need their data type changed, and proceed to change them.
#str(newData)
#Judging from the structure, I opted to change the types of the following:

#Zip_Code from numeric to factor
#lattitude from character to numeric
#longitutde from character to numeric
#Invoice.Item_Number from character to numeric
#Vendor_Number from character to numeric

newData$Zip_Code <- as.factor(newData$Zip_Code)
newData$lattitude <- as.numeric(newData$lattitude)
newData$longitude <- as.numeric(newData$longitude)
newData$Invoice.Item_Number <- as.numeric(newData$Invoice.Item_Number)
newData$Vendor_Number <- as.numeric(newData$Vendor_Number)

#str(newData) checking to see if changes were made

#Finally, I extract the year, month, and day from the Date column and drop it once the new columns are made
#I'm using the same method as above to extract lattitude and longitude from Store_Location

date_split = strsplit(data$Date, split = "/") #this line splits the Date at every instance of a forward slash
newData = transform(newData, Month = sapply(date_split, "[", 1), Day= sapply(date_split, "[", 2), Year = sapply(date_split, "[", 3))

head(newData) #checking if changes were made
#str(newData) checking data types of Month, Day, Year
```
##Part 3: First overview
For the first overview, I'll plot a scatterplot of the lattitude and longitudes of store locations, and then create a visual breakdown of the liquor categories, with volume sold being incorporated as well.
```{r}
library(ggplot2)

ggplot(newData, aes(x=newData$lattitude, y=newData$longitude)) + geom_point() + labs(x = "Lattitude", y= "Longitude", title = "Spread of Liquor Store Locations by Lattitude and Longitude") #initial scatterplot of long. and latt. of the store locations

ggplot(newData, aes(Category_Name, Volume_Sold_.Liters.)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle =90, hjust =1, size = 4)) + labs(title = "Volume of Liquor Sold by Category of Liquor", x = "Category of Liquor", y = "Volume Sold (liters)") #This was the only way I could manage to show the relationship between the category of liquor and the volume sold from each category, if you could comment the solution for this ggplot I would really appreciate it!
```
##Part 4: Data Exploration Cont.
For this part, I found the daily sales liquor sales in Ames in 2018. Then, I had summarized the number of sales, volume of liquor sold and amount of money spent. Must plot volume sold by day (use a scatterplot of volume by day and facet by month), and describe any patterns. Find the dates for ISU football home games in Fall 2018, and look for patterns in liquor purchases.

```{r}
dailySold2018 <- subset(newData, Year =="2018") #subsetting the clean data by year for 2018 sales.

ggplot(dailySold2018, aes(Day, Bottles_Sold)) + geom_point() + facet_wrap(~Month) + labs(title = "Bottles Sold per Day, during Each Month", x = "Day of the Month", y = "Number of Bottles Sold") + theme(axis.text.x = element_text(angle =90, hjust =1, size = 4)) #Total bottles sold per day for each month of 2018

ggplot(dailySold2018, aes(Day, Volume_Sold_.Liters.)) + geom_point() + facet_wrap(~Month) + labs(title = "Volume of Liquor in Liters Sold per Day, during Each Month", x = "Day of the Month", y = "Volume of Liquor (liters)") + theme(axis.text.x = element_text(angle =90, hjust =1, size = 4)) #Total volume of liquor (in liters) sold per day for each month of 2018

ggplot(dailySold2018, aes(Day, Sale_.Dollars.)) + geom_point() + facet_wrap(~Month) + labs(title = "Total Sales in Dollars per Day, during Each Month", x = "Day of the Month", y = "Dollars") + theme(axis.text.x = element_text(angle =90, hjust =1, size = 4)) #Total dollars in sales per day for each month of 2018

```
As far as patterns go, volume of liquor sold is generally higher in the fall semester than in the spring, this could be due to the change in season, as well as the lack of any sports seasons occurring during the spring post-March. I can also see spikes during certain parts of the year. Valentines day, and the week prior to it, had seen a very quick rise in liquor sales. Graduation week in May also had major spikes in liquor sales, but the slow months (as in the volume of liquor sold doesn't exceed 1250 liters on any given day) for liquor sales were as follows: January, March, April, July, August, and November. As far as dates that coincide with ISU home football games go, there is definitely a correlation, with liquor sales being very low, or even zero, on days where the football team wasn't playing. 
On home football days, the volume of liquor sold is higher than on days when the team plays their away games, and it goes to show how much the liquor purchasing habits of those who live in Ames revolve around the sports season.

##Part 5: Dealing with the Date Variable
Use the lubridate package to convert the variable Date from a character to a date. Implement this change.

```{r}
library(lubridate)

str(newData) #Date is a character data type at this point

newData$Date <- mdy(newData$Date) 
str(newData) #the mdy function above changes the data type of Date from character to date

```






